### 2012-03-08 doc/theses/mikula/proposed_algorithm.tex
Master thesis: In progress.

\subsection{Definition of the Syntax Tree}
Firstly, we formally define the syntax tree.

@@

Regarding the additional properties, two constructs in $Q$ represented by two nodes of the same class from $C$ may differ in certain ways, and therefore, it is needed to distinguish them. For instance, two different literal values in $Q$ are represented by nodes $l_1, l_2 \in V$ and $c(l_1) = c(l_2) = LiteralNode$ but each has a different value and type. Therefore $\mathcal{P}$ contains functions \\
$type_{LiteralNode}: V_{LiteralNode} \rightarrow Types_{literal}$ \\
$value_{LiteralNode}: V_{LiteralNode} \rightarrow Values_{literal}$ \\
where $V_{LiteralNode}$ is set $\{v | v \in V, c(v) = LiteralNode\}$, $Types_{literal}$ is set of all types of literal values $\{DECIMAL, INTEGER, DOUBLE, STRING\}$, and $Values_{literal}$ is a set of all literal values (all valid XQuery decimal numbers, integers, double numbers and strings).

@@

For example, an instance of the syntax tree cannot directly contain nodes of $Node$ and $ExprNode$ classes (for every $v \in V, c(v) \neq Node, c(v) \neq ExprNode$), but it can contain nodes of $AttributeNode$ and $LiteralNode$ classes. Regarding the multiplicity of types, a node of $LiteralNode$ class is also considered to be of two indirect types: $ExprNode$ and $Node$.

@@

The node classes can be classified into three groups: inner node classes, leaf node classes and node classes that can be both inner and leaf. Inner node classes (marked $I$) stand for XQuery constructs that are composed of other constructs and can be further divided. An example of such class is $FLWORExprNode$ which is composed of $TupleStreamNode$, $WhereClauseNode$, $OrderByClauseNode$ and $ReturnClauseNode$ classes. Leaf node classes (marked $L$) represent elements of XQuery language that cannot be further divided. For example, $LiteralNode$. Node $FunctionCallNode$ is an example of the third group (marked $IL$). Function call of a function without arguments is represented by a leaf node while function call with arguments is represented by an inner node and its subnodes are nodes representing those arguments.

\begin{figure}
\caption{Syntax tree node types part 1.}
\label{FIG_syntax_tree_node_types_part_1}
\begin{tabular}{l l | l}
\multicolumn{2}{l|}{\textbf{\emph{Node}}} & \textbf{\emph{ExprNode}} \\
\hline
AttributeNode (I) & ModuleNode (I) & CommaOperatorNode (I) \\
AttrListNode (IL) & NameNode (IL) & ConstructorNode (I) \\
AxisNode (I) & OrderByClauseNode (I) & ContextItemExprNode (L) \\
CaseClauseNode (I) & OrderSpecNode (I) & EmptySequenceNode (L) \\
CaseClausesNode (I) & ParamListNode (I) & ExtensionExprNode (I) \\
CDataSectionNode (L) & ParamNode (I) & FLWORExprNode (I) \\
ContentNode (IL) & PITargetNode (IL) & FunctionCallNode (IL) \\
DefaultCaseNode (I) & PragmaListNode (I) & IfExprNode (I) \\
EntityRefNode (L) & PragmaNode (L) & LiteralNode (L) \\
\textbf{ExprHolderNode} & PredicateListNode (I) & OperatorNode (I) \\
\textbf{ExprNode} & \textbf{PrologChildNode} & OrderedExprNode (I) \\
FunctionBodyNode (IL) & StepExprNode (I) & PathExprNode (I) \\
CharRefNode (L) & StringNode (L) & QuantifiedExprNode (I) \\
InClausesNode (I) & TupleStreamNode (I) & TypeswitchExprNode (I) \\
\textbf{ItemTypeNode} & TypeNode (L) & UnorderedExprNode (I) \\
LocationHintNode (L) & \textbf{VariableBindingNode} & ValidateExprNode (I) \\
LocationHintsNode (IL) & VarValueNode (IL) & VarRefNode (L) \\
\textbf{ModuleChildNode} & & \\
\end{tabular}
\end{figure}

\begin{figure}
\caption{Syntax tree node types part 2.}
\label{FIG_syntax_tree_node_types_part_2}
\begin{tabular}{l | l}
\textbf{\emph{PrologChildNode}} & \textbf{\emph{ExprHolderNode}} \\
\hline
BaseURIDeclNode (L) & BindingSequenceNode (I) \\
BoundarySpaceDeclNode (L) & DefaultValueNode (I) \\
ConstructionDeclNode (L) & ElseExpressionNode (I) \\
CopyNamespacesDeclNode (L) & OperandExpressionNode (I) \\
DefaultCollationDeclNode (L) & ReturnClauseNode (I) \\
DefaultNamespaceDeclNode (L) & TestExpressionNode (I) \\
EmptyOrderDeclNode (L) & ThenExpressionNode (I) \\
FunctionDeclNode (I) & WhereClauseNode (I) \\
\cline{2-2} \textbf{ImportNode} & \textbf{\emph{ItemTypeNode}} \\
\cline{2-2} NamespaceDeclNode (L) & AnyItemNode (L) \\
OptionDeclNode (L) & AtomicTypeNode (L) \\
OrderingModeDeclNode (L) & KindTestNode (IL) \\
VarDeclNode (I) & NameTestNode (L) \\
\cline{1-2} \textbf{\emph{ImportNode}} & \textbf{\emph{VariableBindingNode}} \\
\cline{1-2} ModuleImportNode (I) & ForClauseNode (I) \\
SchemaImportNode (I) & InClauseNode (I) \\
\cline{1-1} \textbf{\emph{ModuleChildNode}} & LetClauseNode (I) \\
\cline{1-2} ModuleDeclNode (L) & \emph{StepExprNode} (I) \\
\cline{2-2}PrologNode (IL) & SelfOrDescendantStepNode (L) \\
QueryBodyNode (I) & 
\end{tabular}
\end{figure}

@@

\begin{define}[Function is(v, class)]For every $v \in V$ and every $class \in C$, function $is(v, class)$ returns boolean value $true$ if $c(v) = class$ or $c(v)$ is a (direct or indirect) subclass of $class$, according to the described principle. Otherwise, it returns $false$.
\end{define}

For instance, assuming $v \in V, c(v) = LiteralNode$, calls $is(v, LiteralNode)$, $is(v, ExprNode)$ returns $true$, while call $is(v, ContentNode)$ returns $false$.

\subsection{Characteristics of the Syntax Tree}
An important characteristic of the syntax tree is related to a definition of local variables and their scope in the XQuery language. The representation of a definition of a local variable in the syntax tree is a node of $VariableBinding$ class. Nodes of that class have only two children; a node representing the type of the variable and a node representing the binding expression (expression defining the value of the variable, and thus, it cannot use the variable). Hence, the entire subtree does not contain any expressions that use the variable. Therefore the scope of the new variable is not the subtree of the $VariableBinding$ class node. It depends on the type of XQuery construct that the variable binding is a (indirect) component of.

For example, a node of $FLWORExpr$ class contains four subnodes of\linebreak $TupleStream$, $WhereClause$, $OrderByClause$ and $ReturnClause$ classes. The $TupleStream$ class node contains a list of node of $variableBinding$ class which define variables valid in all other three subnodes of the FLWOR node.

@@

\begin{itemize}
\item XML Schema built-in atomic types \todo{link do appendixu}.
\item Types $ElementType$, $AttributeType$, $NodeType$, $TextNodeType$,\linebreak $CommentType$, $ProcessingInstructionType$, $DocumentType$ representing an element, attribute, node, text node, comment, processing instruction, document node.
\item Type representing a node or a set of nodes selected by a certain path expression. The path expression is included in this type. Let this type be identified as $PathType$.
\item $UnknownType$ representing a type without known details, which does not suit one of the three previous types. An example is XSD type $anyType$.
\end{itemize}

\subsection{PathType}
$PathType$ contains additional information. The represented path is contained by an instance of $PathExprNode$ containing list of its steps, in particular instances of $StepExprNode$. If a step is a reference to a variable which type is $PathType$, we also want to include this information. Therefore, $PathType$ contain association between the steps and other $PathTypes$ and this association is defined for the $PathType$ variable steps.

To differ between a common $PathType$ selecting a set of nodes and a $PathType$ bound to a for variable in a FLWOR expression, $PathType$ structure contains a flag $isForBound$.

Additionaly, $PathType$ contains a list of special functions that were called with an argument of $PathType$ type. The motivation is that in some cases of function calls, we want to know that the function call is performed with an instance of $PathType$ because then, we can determine a type of the function call more precisely. Those special functions are built-in functions $data$, $min$, $max$ \todo{Doplnit}.

In summary, we represent $PathType$ as a structure with the following member variables. \todo{Radsej UML?}
\begin{itemize}
\item $pathExprNode$ - An instance of $PathExprNode$.
\item $substeps$ - An association between references from $stepNodes$ and instances of $PathType$ type.
\item $isForBound$ - Boolean flag determining if the type was bound to a for variable in a for clause.
\item $specialFunctionCalls$ - List of special functions called with this instance as an argument.
\end{itemize}

@@

To capture sequences, we assign the first two categories of types (all types except of $PathType$ and $UnknownType$) with its cardinality as proposed in \todo{ref na Schajbalovu diplomku}. Each of those types can be perceived as a sequence. A type representing one value or one item can be perceived as a sequence of exactly one item. The cardinality expresses one of the following five sequence types.

\begin{itemize}
\item An empty sequence.
\item A sequence of exactly one item.
\item A sequence containing zero or one item (modifier $?$).
\item A sequence containing zero or more items (modifier $*$).
\item A sequence containing one or more items (modifier $+$).
\end{itemize}

$PathType$ is not assigned with the cardinality since we do not evaluate the queries, and therefore, we cannot determine if a certain XQuery path targets zero, one or more nodes. Alike, $UnknownType$ is neither assigned with the cardinality. Expressions of $UnknownType$ are not utilized it the inference, therefore, their cardinality is not needed.

@@

\begin{algorithm}
\caption{Retrieval of Function Declaration Nodes $getFunctionDeclarationNodes(syntaxTree)$}
\label{ALG_retrieval_of_function_declaration_nodes}
\begin{algorithmic}[1]
\REQUIRE $syntaxTree$: A reference to the root node of a syntax tree.
\ENSURE A list of syntax tree nodes representing function declarations.

\STATE $prologNode = null$
\FORALL{$moduleChildNode \in syntaxTree.getChildren()$}
    \IF{$is(moduleChildNode, PrologNode)$}
        \STATE $prologNode = moduleChildNode$
    \ENDIF
\ENDFOR

\STATE $functionDeclarationNodes =$ an empty list
\IF{$prologNode \neq null$}
    \FORALL{$prologChildNode \in prologNode.getChildren()$}
        \IF{$is(prologChildNode, FunctionDeclNode)$}
            \STATE add $prologChildNode$ to $functionDeclarationNodes$
        \ENDIF
    \ENDFOR
\ENDIF

\RETURN $functionDeclarationNodes$
\end{algorithmic}
\end{algorithm}

A simple algorithm is presented in Algorithm \ref{ALG_processing_of_functions}. It uses \texttt{getFunction\-DeclarationNodes()} function, defined in Algorithm \ref{ALG_retrieval_of_function_declaration_nodes}, which returns a list of all function declaration nodes in the syntax three. Actually, it does not have to search the whole syntax tree as the function declaration nodes can be present only in the query prolog section.

Without loss of generality, we will focus on locally defined functions. We will not determine types of functions defined in other modules since the principle is similar but it is needed to look up the definitions in syntax trees of other queries. Return types of built-in functions are fixed and, thus, there is no need to analyze them. Also, since a determination of a prefix for built-in functions is a technical issue, we assume that built-in functions are either prefixed by $fn$ or not prefixed.

\begin{algorithm}
\caption{Processing of Functions}
\label{ALG_processing_of_functions}
\begin{algorithmic}[1]
\REQUIRE $syntaxTree$: A reference to the root node of a syntax tree.
\ENSURE An associative array of function names with their types and references to the their declaration nodes.

\STATE $functionArray$ = an empty associative array
\FORALL{$functionDeclarationNode \in$ \\ $getFunctionDeclarationNodes(syntaxTree)$}
    \STATE $functionName = functionDeclarationNode.funcName$
    \STATE $typeNode = functionDeclarationNode.getChild(TypeNode)$
    \STATE $functionArray[functionName] =$ \\ $(getTypeTN(typeNode), functionDeclarationNode)$
\ENDFOR
\RETURN $functionArray$
\end{algorithmic}
\end{algorithm}

\begin{algorithm}
\caption{Extraction of a Type from TypeNode getTypeTN()}
\label{ALG_extraction_of_a_type_from_typenode}
\begin{algorithmic}[1]
\REQUIRE $typeNode$: Syntax tree node of TypeNode class.
\ENSURE Type extracted from $typeNode$.

\STATE $type = null$
\STATE $cardinality = typeNode.cardinality$
\STATE $itemTypeNode = typeNode.getChild(ItemTypeNode)$
\IF{$is(itemTypeNode, AtomicTypeNode)$}
    \STATE $type = itemTypeNode.typeName$
\ELSIF{$is(itemTypeNode, KindTestNode)$}
    \STATE $type = itemTypeNode.nodeKind$
\ENDIF

\RETURN $type$, $cardinality$
\end{algorithmic}
\end{algorithm}

@@

For the rest of the thesis, we assume function $getFunctionReturnType()$ which takes a function name and returns the return type of the function if the function is either built-in or it is recorded in the associative array. Otherwise, \texttt{null} is returned.

Later phases of the algorithm need to process the declarations of certain functions. Therefore, we also store references to the entire function declaration nodes, accessible using function $getFunctionDeclNode()$.

@@

Function $memorizeType()$ takes two arguments $node$ and $type$ and it memorizes the given type of the specified expression node. This information can be then retrieved by function $getType()$, specifying the particular node as its argument.

Another group consists of functions $set()$, $add()$, and $get()$. Function $set()$ memorizes a given value of a specified property of a specified node. It is used to assign a node with a named value, for example, $set(someNode, color, 'red')$ will assign $someNode$ with string value $'red'$ wich can be then retrieved by function $get()$, specifying the node and the property name. For example,\linebreak $get(someNode, color)$ returns $'red'$. A subsequent call of $set()$ assigning a node with a value of already assigned property will overwrite it, leaving the property with the newer value. However, a property can have several values and this can be achieved using function $add()$ with the same syntax as function $set()$ but instead of overwriting the existing property, $add()$ will append the new value to the existing ones. Then, a call of $get()$ on this node and property returns a list of all its values, preserving the order of their addition.

@@

\begin{algorithm}
\caption{Function determineExpressionType()}
\label{ALG_func_determineExpressionType}
\begin{algorithmic}[1]
\REQUIRE{\ \\
    $exprNode$: An expression node in the syntax tree. \\
    $contextVarTypes$: Types of local variables valid in the current subtree.
}
\ENSURE The expression type.

\STATE $type = UnknownType$
\IF{$is(exprNode, LiteralNode)$}
     \STATE $type =$ XSD atomic type $exprNode.type$, exactly-once cardinality
\ELSIF{$is(exprNode, FunctionCallNode)$}
     \STATE $type = getFunctionReturnType(exprNode.fncName)$
\ELSIF{$is(exprNode, VarRefNode)$}
     \STATE $type = getVariableType(contextVarTypes, exprNode.varName)$
     \IF{$type$ is $PathType$}
         \STATE $step$ = a path step (new instance of $StepExprNode$) representing the variable reference
		 \STATE $substeps$ = empty associative array         
         \STATE $substeps[step]$ = $type$
         \STATE $pathExprNode$ = new instance of $PathExprNode$ containing one step $step$
         \STATE $type$ = $PathType$ containing $pathExprNode$ and $substeps$
     \ENDIF
\ELSIF{$is(exprNode, PathExprNode)$}
     \STATE $type = createPathType(exprNode)$
\ELSIF{$is(exprNode, OperatorNode)$}
     \STATE $type = determineOperatorType(exprNode, contextVarTypes)$
\ELSIF{$is(exprNode, FLWORExprNode)$}
     \STATE $returnClauseNode = exprNode.getChild(ReturnClauseNode)$
     \STATE $type = createForUnboundType(getType(returnClauseNode))$
\ENDIF

\RETURN $type$
\end{algorithmic}
\end{algorithm}

\begin{algorithm}
\caption{Function determineOperatorType()}
\label{ALG_func_determineOperatorType}
\begin{algorithmic}[1]
\REQUIRE{\ \\ 
    $operatorNode$: An operator expression node in the syntax tree. \\
    $contextVarTypes$: Types of local variables valid in the current subtree.
}
\ENSURE The operator expression type.

>>>>

\section{Definition of the Syntax Tree}
Firstly, we formally define the syntax tree.

@@

Regarding the additional properties, two constructs in $Q$ represented by two nodes of the same class from $C$ may differ in certain ways, and therefore, it is needed to distinguish them. For instance, two different literal values in $Q$ are represented by nodes $l_1, l_2 \in V$ and $c(l_1) = c(l_2) =$ \texttt{LiteralNode} but each has a different value and type. Therefore $\mathcal{P}$ contains functions \\
$type_{LiteralNode}: V_{LiteralNode} \rightarrow Types_{literal}$ \\
$value_{LiteralNode}: V_{LiteralNode} \rightarrow Values_{literal}$ \\
where $V_{LiteralNode}$ is set $\{v | v \in V, c(v) =$ \texttt{LiteralNode}$\}$, $Types_{literal}$ is set of all types of literal values \{\texttt{DECIMAL}, \texttt{INTEGER}, \texttt{DOUBLE}, \texttt{STRING}\}, and $Values_{literal}$ is a set of all literal values (all valid XQuery decimal numbers, integers, double numbers and strings).

@@

For example, an instance of the syntax tree cannot directly contain nodes of \texttt{Node} and \texttt{ExprNode} classes (for every $v \in V, c(v) \neq$ \texttt{Node}$, c(v) \neq$ \texttt{ExprNode}), but it can contain nodes of \texttt{AttributeNode} and \texttt{LiteralNode} classes. Regarding the multiplicity of types, a node of \texttt{LiteralNode} class is also considered to be of two indirect types: \texttt{ExprNode} and \texttt{Node}.

@@

The node classes can be classified into three groups: inner node classes, leaf node classes and node classes that can be both inner and leaf. Inner node classes (marked \texttt{I}) stand for XQuery constructs that are composed of other constructs and can be further divided. An example of such class is \texttt{FLWORExprNode} which is composed of \texttt{TupleStreamNode}, \texttt{WhereClauseNode}, \texttt{OrderByClauseNode} and \texttt{ReturnClauseNode} classes. Leaf node classes (marked \texttt{L}) represent elements of XQuery language that cannot be further divided. For example, \texttt{LiteralNode}. Node \texttt{FunctionCallNode} is an example of the third group (marked \texttt{IL}). Function call of a function without arguments is represented by a leaf node while function call with arguments is represented by an inner node and its subnodes are nodes representing those arguments.

\begin{figure}
\caption{Syntax tree node types part 1.}
\label{FIG_syntax_tree_node_types_part_1}
\begin{tabular}{l l | l}
\multicolumn{2}{l|}{\textbf{\emph{Node}}} & \textbf{\emph{ExprNode}} \\
\hline
\texttt{AttributeNode (I)} & \texttt{ModuleNode (I)} & \texttt{CommaOperatorNode (I)} \\
\texttt{AttrListNode (IL)} & \texttt{NameNode (IL)} & \texttt{ConstructorNode (I)} \\
\texttt{AxisNode (I)} & \texttt{OrderByClauseNode (I)} & \texttt{ContextItemExprNode (L)} \\
\texttt{CaseClauseNode (I)} & \texttt{OrderSpecNode (I)} & \texttt{EmptySequenceNode (L)} \\
\texttt{CaseClausesNode (I)} & \texttt{ParamListNode (I)} & \texttt{ExtensionExprNode (I)} \\
\texttt{CDataSectionNode (L)} & \texttt{ParamNode (I)} & \texttt{FLWORExprNode (I)} \\
\texttt{ContentNode (IL)} & \texttt{PITargetNode (IL)} & \texttt{FunctionCallNode (IL)} \\
\texttt{DefaultCaseNode (I)} & \texttt{PragmaListNode (I)} & \texttt{IfExprNode (I)} \\
\texttt{EntityRefNode (L)} & \texttt{PragmaNode (L)} & \texttt{LiteralNode (L)} \\
\textbf{ExprHolderNode} & \texttt{PredicateListNode (I)} & \texttt{OperatorNode (I)} \\
\textbf{ExprNode} & \textbf{PrologChildNode} & \texttt{OrderedExprNode (I)} \\
\texttt{FunctionBodyNode (IL)} & \texttt{StepExprNode (I)} & \texttt{PathExprNode (I)} \\
\texttt{CharRefNode (L)} & \texttt{StringNode (L)} & \texttt{QuantifiedExprNode (I)} \\
\texttt{InClausesNode (I)} & \texttt{TupleStreamNode (I)} & \texttt{TypeswitchExprNode (I)} \\
\textbf{ItemTypeNode} & \texttt{TypeNode (L)} & \texttt{UnorderedExprNode (I)} \\
\texttt{LocationHintNode (L)} & \textbf{VariableBindingNode} & \texttt{ValidateExprNode (I)} \\
\texttt{LocationHintsNode (IL)} & \texttt{VarValueNode (IL)} & \texttt{VarRefNode (L)} \\
\textbf{ModuleChildNode} & & \\
\end{tabular}
\end{figure}

\begin{figure}
\caption{Syntax tree node types part 2.}
\label{FIG_syntax_tree_node_types_part_2}
\begin{tabular}{l | l}
\textbf{\emph{PrologChildNode}} & \textbf{\emph{ExprHolderNode}} \\
\hline
\texttt{BaseURIDeclNode (L)} & \texttt{BindingSequenceNode (I)} \\
\texttt{BoundarySpaceDeclNode (L)} & \texttt{DefaultValueNode (I)} \\
\texttt{ConstructionDeclNode (L)} & \texttt{ElseExpressionNode (I)} \\
\texttt{CopyNamespacesDeclNode (L)} & \texttt{OperandExpressionNode (I)} \\
\texttt{DefaultCollationDeclNode (L)} & \texttt{ReturnClauseNode (I)} \\
\texttt{DefaultNamespaceDeclNode (L)} & \texttt{TestExpressionNode (I)} \\
\texttt{EmptyOrderDeclNode (L)} & \texttt{ThenExpressionNode (I)} \\
\texttt{FunctionDeclNode (I)} & \texttt{WhereClauseNode (I)} \\
\cline{2-2} \textbf{ImportNode} & \textbf{\emph{ItemTypeNode}} \\
\cline{2-2} \texttt{NamespaceDeclNode (L)} & \texttt{AnyItemNode (L)} \\
\texttt{OptionDeclNode (L)} & \texttt{AtomicTypeNode (L)} \\
\texttt{OrderingModeDeclNode (L)} & \texttt{KindTestNode (IL)} \\
\texttt{VarDeclNode (I)} & \texttt{NameTestNode (L)} \\
\cline{1-2} \textbf{\emph{ImportNode}} & \textbf{\emph{VariableBindingNode}} \\
\cline{1-2} \texttt{ModuleImportNode (I)} & \texttt{ForClauseNode (I)} \\
\texttt{SchemaImportNode (I)} & \texttt{InClauseNode (I)} \\
\cline{1-1} \textbf{\emph{ModuleChildNode}} & \texttt{LetClauseNode (I)} \\
\cline{1-2} \texttt{ModuleDeclNode (L)} & \texttt{\emph{StepExprNode} (I)} \\
\cline{2-2} \texttt{PrologNode (IL)} & \texttt{SelfOrDescendantStepNode (L)} \\
\texttt{QueryBodyNode (I)} & 
\end{tabular}
\end{figure}

@@

\begin{define}[Function $is(v, class)$]For every $v \in V$ and every $class \in C$, function $is(v, class)$ returns boolean value $true$ if $c(v) = class$ or $c(v)$ is a (direct or indirect) subclass of $class$, according to the described principle. Otherwise, it returns $false$.
\end{define}

For instance, assuming $v \in V, c(v) =$ \texttt{LiteralNode}, calls $is(v,$ \texttt{LiteralNode}$)$ and $is(v,$ \texttt{ExprNode}$)$ return $true$, while call $is(v,$ \texttt{ContentNode}$)$ returns $false$.

\subsection{Characteristics of the Syntax Tree}
An important characteristic of the syntax tree is related to a definition of local variables and their scope in the XQuery language. The representation of a definition of a local variable in the syntax tree is a node of \texttt{VariableBindingNode} class. Nodes of that class have only two children; a node representing the type of the variable and a node representing the binding expression (expression defining the value of the variable, and thus, it cannot use the variable). Hence, the entire subtree does not contain any expressions that use the variable. Therefore the scope of the new variable is not the subtree of the \texttt{VariableBindingNode} class node. It depends on the type of XQuery construct that the variable binding is a (indirect) component of.

For example, a node of \texttt{FLWORExprNode} class contains four subnodes of\linebreak \texttt{TupleStreamNode}, \texttt{WhereClauseNode}, \texttt{OrderByClauseNode} and\linebreak \texttt{ReturnClauseNode} classes. The \texttt{TupleStreamNode} class node contains a list of nodes of \texttt{VariableBindingNode} class which define variables valid in all other three subnodes of the FLWOR node.

@@

\begin{itemize}
\item XML Schema built-in atomic types \todo{link do appendixu}.
\item Types \texttt{ElementType}, \texttt{AttributeType}, \texttt{NodeType}, \texttt{TextNodeType},\linebreak \texttt{CommentType}, \texttt{ProcessingInstructionType}, \texttt{DocumentType} representing an element, attribute, node, text node, comment, processing instruction, document node.
\item Type representing a node or a set of nodes selected by a certain path expression. The path expression is included in this type. Let this type be identified as \texttt{PathType}.
\item \texttt{UnknownType} representing a type without known details, which does not suit one of the three previous types. An example is XSD type \texttt{anyType}.
\end{itemize}

\subsection{PathType}
\texttt{PathType} contains additional information. The represented path is contained by an instance of \texttt{PathExprNode} containing list of its steps, in particular instances of \texttt{StepExprNode}. If a step is a reference to a variable which type is \texttt{PathType}, we also want to include this information. Therefore, \texttt{PathType} contain association between the steps and other \texttt{PathType}s and this association is defined for the \texttt{PathType} variable steps.

To differ between a common \texttt{PathType} selecting a set of nodes and a \texttt{PathType} bound to a for variable in a FLWOR expression, \texttt{PathType} structure contains a flag \texttt{isForBound}.

Additionally, \texttt{PathType} contains a list of special functions that were called with an argument of \texttt{PathType} type. The motivation is that in some cases of function calls, we want to know that the function call is performed with an instance of \texttt{PathType} because then, we can determine a type of the function call more precisely. Those special functions are built-in functions \texttt{data}, \texttt{min}, \texttt{max} \todo{Doplnit}.

In summary, we represent \texttt{PathType} as a structure with the following member variables.
\begin{itemize}
\item \texttt{pathExprNode} - An instance of \texttt{PathExprNode}.
\item \texttt{substeps} - An association between steps from \texttt{pathExprNode} and instances of \texttt{PathType} type.
\item \texttt{isForBound} - Boolean flag determining if the type was bound to a for variable in a for clause.
\item \texttt{specialFunctionCalls} - List of special functions called with this instance as an argument.
\end{itemize}

@@

To capture sequences, we assign the first two categories of types (all types except of \texttt{PathType} and \texttt{UnknownType}) with its cardinality as proposed in \todo{ref na Schajbalovu diplomku}. Each of those types can be perceived as a sequence. A type representing one value or one item can be perceived as a sequence of exactly one item. The cardinality expresses one of the following five sequence types.

\begin{itemize}
\item An empty sequence.
\item A sequence of exactly one item.
\item A sequence containing zero or one item (modifier \texttt{?}).
\item A sequence containing zero or more items (modifier \texttt{*}).
\item A sequence containing one or more items (modifier \texttt{+}).
\end{itemize}

\texttt{PathType} is not assigned with the cardinality since we do not evaluate the queries, and therefore, we cannot determine if a certain XQuery path targets zero, one or more nodes. Alike, \texttt{UnknownType} is neither assigned with the cardinality. Expressions of \texttt{UnknownType} are not utilized it the inference, therefore, their cardinality is not needed.

@@

\begin{algorithm}
\caption{Function \texttt{getFunctionDeclarationNodes}: Retrieval of Function Declaration Nodes}
\label{ALG_retrieval_of_function_declaration_nodes}
\begin{algorithmic}[1]
\REQUIRE syntaxTree: A reference to the root node of a syntax tree.
\ENSURE A list of syntax tree nodes representing function declarations.

\STATE prologNode $=$ null
\FORALL{moduleChildNode $\in$ syntaxTree.getChildren()}
    \IF{is(moduleChildNode, PrologNode)}
        \STATE prologNode $=$ moduleChildNode
    \ENDIF
\ENDFOR

\STATE functionDeclarationNodes $=$ an empty list
\IF{prologNode $\neq$ null}
    \FORALL{prologChildNode $\in$ prologNode.getChildren()}
        \IF{is(prologChildNode, FunctionDeclNode)}
            \STATE add prologChildNode to functionDeclarationNodes
        \ENDIF
    \ENDFOR
\ENDIF

\RETURN functionDeclarationNodes
\end{algorithmic}
\end{algorithm}

A simple algorithm is presented in Algorithm \ref{ALG_processing_of_functions}. It uses\linebreak \texttt{getFunctionDeclarationNodes} function, defined in Algorithm \ref{ALG_retrieval_of_function_declaration_nodes}, which returns a list of all function declaration nodes in the syntax three. Actually, it does not have to search the whole syntax tree as the function declaration nodes can be present only in the query prolog section.

Without loss of generality, we will focus on locally defined functions. We will not determine types of functions defined in other modules since the principle is similar but it is needed to look up the definitions in syntax trees of other queries. Return types of built-in functions are fixed and, thus, there is no need to analyse them. Also, since a determination of a prefix for built-in functions is a technical issue, we assume that built-in functions are either prefixed by $fn$ or not prefixed.

\begin{algorithm}
\caption{Processing of Functions}
\label{ALG_processing_of_functions}
\begin{algorithmic}[1]
\REQUIRE syntaxTree: A reference to the root node of a syntax tree.
\ENSURE An associative array of function names with their types and references to the their declaration nodes.

\STATE functionArray $=$ an empty associative array
\FORALL{functionDeclarationNode $\in$ \\ getFunctionDeclarationNodes(syntaxTree)}
    \STATE functionName $=$ functionDeclarationNode.funcName
    \STATE typeNode $=$ functionDeclarationNode.getChild(TypeNode)
    \STATE functionArray[functionName] $=$ \\ (getTypeTN(typeNode), functionDeclarationNode)
\ENDFOR
\RETURN functionArray
\end{algorithmic}
\end{algorithm}

\begin{algorithm}
\caption{Function \texttt{getTypeTN}: Extraction of a Type from TypeNode}
\label{ALG_extraction_of_a_type_from_typenode}
\begin{algorithmic}[1]
\REQUIRE typeNode: Syntax tree node of TypeNode class.
\ENSURE Type extracted from typeNode.

\STATE type = null
\STATE cardinality = typeNode.cardinality
\STATE itemTypeNode = typeNode.getChild(ItemTypeNode)
\IF{is(itemTypeNode, AtomicTypeNode)}
    \STATE type = itemTypeNode.typeName
\ELSIF{is(itemTypeNode, KindTestNode)}
    \STATE type = itemTypeNode.nodeKind
\ENDIF

\RETURN type, cardinality
\end{algorithmic}
\end{algorithm}

@@

For the rest of the thesis, we assume function \texttt{getFunctionReturnType} which takes a function name and returns the return type of the function if the function is either built-in or it is recorded in the associative array. Otherwise, \texttt{null} is returned.

Later phases of the algorithm need to process the declarations of certain functions. Therefore, we also store references to the entire function declaration nodes, accessible using function \texttt{getFunctionDeclNode}.

@@

Function \texttt{memorizeType} takes two arguments \texttt{node} and \texttt{type} and it memorizes the given type of the specified expression node. This information can be then retrieved by function \texttt{getType}, specifying the particular node as its argument.

Another group consists of functions \texttt{set}, \texttt{add}, and \texttt{get}. Function \texttt{set} memorizes a given value of a specified property of a specified node. It is used to assign a node with a named value, for example, \texttt{set(someNode, color, "red")} will assign \texttt{someNode} with string value \texttt{"red"} which can be then retrieved by function \texttt{get}, specifying the node and the property name. For example,\linebreak \texttt{get(someNode, color)} returns \texttt{"red"}. A subsequent call of \texttt{set} assigning a node with a value of already assigned property will overwrite it, leaving the property with the newer value. However, a property can have several values and this can be achieved using function \texttt{add} with the same syntax as function \texttt{set} but instead of overwriting the existing property, \texttt{add} will append the new value to the existing ones. Then, a call of \texttt{get} on this node and property returns a list of all its values, preserving the order of their addition.

@@

\begin{algorithm}
\caption{Function \texttt{determineExpressionType}}
\label{ALG_func_determineExpressionType}
\begin{algorithmic}[1]
\REQUIRE{\ \\
    exprNode: An expression node in the syntax tree. \\
    contextVarTypes: Types of local variables valid in the current subtree.
}
\ENSURE The expression type.

\STATE type $=$ UnknownType
\IF{is(exprNode, LiteralNode)}
     \STATE type $=$ XSD atomic type exprNode.type, exactly-once cardinality
\ELSIF{is(exprNode, FunctionCallNode)}
     \STATE type $=$ getFunctionReturnType(exprNode.fncName)
\ELSIF{is(exprNode, VarRefNode)}
     \STATE type = getVariableType(contextVarTypes, exprNode.varName)
     \IF{type is PathType}
         \STATE step $=$ a path step (new instance of StepExprNode) representing the variable reference
		 \STATE substeps = empty associative array         
         \STATE substeps[step] $=$ type
         \STATE pathExprNode $=$ new instance of PathExprNode containing one step step
         \STATE type $=$ PathType containing pathExprNode and substeps
     \ENDIF
\ELSIF{is(exprNode, PathExprNode)}
     \STATE type $=$ createPathType(exprNode)
\ELSIF{is(exprNode, OperatorNode)}
     \STATE type $=$ determineOperatorType(exprNode, contextVarTypes)
\ELSIF{is(exprNode, FLWORExprNode)}
     \STATE returnClauseNode $=$ exprNode.getChild(ReturnClauseNode)
     \STATE type $=$ createForUnboundType(getType(returnClauseNode))
\ENDIF

\RETURN type
\end{algorithmic}
\end{algorithm}

\begin{algorithm}
\caption{Function \texttt{determineOperatorType}}
\label{ALG_func_determineOperatorType}
\begin{algorithmic}[1]
\REQUIRE{\ \\ 
    $operatorNode$: An operator expression node in the syntax tree. \\
    $contextVarTypes$: Types of local variables valid in the current subtree.
}
\ENSURE The operator expression type.

\STATE $type =$ UnknownType
\STATE $operator =$ $operatorNode.operator$
\IF{isOperatorClassComparison($operator$)}
    \STATE $type =$ boolean, exactly-one cardinality
\ELSIF{isOperatorClassAddition($operator$)}
    \STATE $leftOperandType =$ getType($exprNode.leftSide$)
    \STATE $rightOperandType =$ getType($exprNode.rightSide$)
    \IF{isNumericType($leftOperandType$) $\wedge$ isNumericType($rightOperandType$)}
        \STATE $type =$ selectMoreGeneralNumericType($leftOperandType$, $rightOperandType$)
    \ELSIF{isNumericType($leftOperandType$)}
        \STATE $type = leftOperandType$
    \ELSIF{isNumericType($rightOperandType$)}
        \STATE $type = rightOperandType$
    \ENDIF
\ENDIF

\RETURN type
\end{algorithmic}
\end{algorithm}

\begin{algorithm}
\caption{Function \texttt{isOperatorClassComparison}}
\label{ALG_func_isOperatorClassComparison}
\begin{algorithmic}[1]
\REQUIRE $operator$: Representation of a XQuery operator.
\ENSURE true if operator is a comparison operator, false otherwise.
